Handling String-Number Conversion in String Constraint Solving

### 2019-11-05 draft.tex
some bug fix

@@

@@

A \emph{finite automaton} is a tuple $(Q,T,\Sigma,q_0,q_m)$, where $Q$ is the set of states, $T\subseteq Q\times (\Sigma\cup \{\epsilon\} )\times Q $ is the set of transition relation, $\Sigma$ is the alphabet, $q_0$ is the initial state, and $q_m$ is the final state. The semantics of finite automaton is defined in the standard manner. We use $L(A)$ to denote the regular language, i.e., set of accepted strings, of a finite automaton $A$.

@@

Given a finite automaton $A$ over $\{a_1,a_2,\ldots,a_n\}$, the Parikh image of $L$ is the set $\{(|w|_{a_1},\ldots, |w|_{a_n}) \mid w \in  L(A) \}$. Moreover, one can construct from $A$ in linear time a Presburger formula $\phi_A(v_1, \ldots,v_n)$ such that $\phi(c_1, \ldots,c_n) \iff (c_1, \ldots,c_n)$ in the Parikh image of $L$~\cite{peter}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Over-Approximation and Under-Approximation Framework} \label{section:framework}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For proving UNSAT, our strategy first over-approximates string-number conversion function $n=\its{x}$ as an uninterpreted function with an additional axiom $n\leq 0 \rightarrow \sti{\its{n}}=n$.

>>>>

String constraint solving has received considerable attention in the constraint solving community, with solvers such as CVC4 and Z3 implmenting solvers of ever-increasing sophistication.  Operations such as substring and string length are widely supported, and recent solvers such as Z3-str3 have shown impressive results.  However, these solvers have limited for operations that convert between strings and integers; string length operations are sometimes well supported, but not parsing a string as an integer or turning an integer value into its string form.  When there is support, it is often limited to ground terms, i.e. at least one side of the conversion must be a constant.  Thus, $v = string2int("5")$ or $v = int2string(5)$ are handled but $v_1 = int2string(v_2)$ will fail.

As solvers such Z3 are applied to symbolic execution of programming languages, this is increasingly becoming a severe limitation.  For many languages such as Java, conversion between integers and strings is a niche operation, and much symbolic execution can be done without supporting it; however, this is not true of some scripting languages, of which JavaScript is the most notable.  JavaScript is a key language---it powers most interactive content on the Web, and increasingly server-side code with nodejs---and it has conversions between strings and integers embedded in the core of its semantics.

\begin{verbatim}
function resetArray(arr) {
  for(var i = 0; i < arr.length; i++) {
    arr[i] = 0
  }
}
\end{verbatim}

A casual glance at the above code reveal no use of strings at all, but the semantics of field access is somewhat unusual in JavaScript: all array indices act like strings, and numeric indices are converted to strings.  Any faithful symbolic execution of JavaScript must handle such conversions for even basic array operations to work correctly; consider the following code snippet that uses {\tt{resetArray}}, with the value of {\tt{x}} shown on the right:

\begin{tabular}{l|c}
	{\tt{x = [0, 1, 2, 3, 4, 5]}} & [0, 1, 2, 3, 4, 5] \\
	{\tt{resetArray(x)}} & [0, 0, 0, 0, 0, 0] \\
	{\tt{x["3"] = 5}} & [0, 0, 0, 5, 0, 0] \\
	{\tt{x["12"] = 7}} & [0, 0, 0, 5, 0, 0, , , , , , 7] \\
	{\tt{resetArray(x)}} & [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0] \\
\end{tabular}

In this code, string and integer values access the same array elements interchangeably, necessitating conversions between string and integers to faithfully model JavaScript semantics.  This conversion is mandated explicitly by JavaScript semantics: the 2019 edition of ECMAScript requires that $ToPropertyKey$ be called on the element expression (\S{12.3.2.1}), and $ToPropertyKey$ calls {\tt{ToString}} on that value in all but special cases (\S{7.1.14}).

Now observe that {\tt{resetArray}} assigns to the array in a loop of which the size cannot, in general, be known a priori; as such, falls into the case where neither the string nor the integer in a conversion are constant and hence most SMT solvers will not be able to handle it.  But this is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code.

\todo{The end of Julian's current writing}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview} \label{section:overview}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For proving UNSAT, our strategy first over-approximates string-number conversion function $n=\its{x}$ as an uninterpreted function with an additional axiom $n\leq 0 \rightarrow \sti{\its{n}}=n$.

@@

A \emph{finite automaton} is a tuple $(Q,T,\Sigma,q_0,q_m)$, where $Q$ is the set of states, $T\subseteq Q\times (\Sigma\cup \{\epsilon\} )\times Q $ is the set of transition relation, $\Sigma$ is the alphabet, $q_0$ is the initial state, and $q_m$ is the final state. The semantics of finite automaton is defined in the standard manner. We use $L(A)$ to denote the regular language, i.e., set of accepted strings, of the automaton $A$.

@@

Given a finite automaton $A$ over $\{a_1,a_2,\ldots,a_n\}$, the Parikh image of $L$ is the set $\{(|w|_{a_1},\ldots, |w|_{a_n}) \mid w \in  L(A) \}$, i.e., the number of occurrences of each symbol $\{a_1,a_2,\ldots,a_n\}$ of words in $L(A)$. Moreover, one can construct from $A$ in linear time a Presburger formula $\phi_A(v_1, \ldots,v_n)$ such that $\phi(c_1, \ldots,c_n) \iff (c_1, \ldots,c_n)$ in the Parikh image of $L$~\cite{peter}.

@@
